Learning features and abstract actions for computing generalized plans

### 2018-09-05 doc/paper.tex
Draft sent to H&B

\begin{itemize}
\item $H: \abs{holding}$: whether any block is being held,
\item $X: \abs{holding \sqcap \set{x}}$: whether block $x$ is being held,
\item $n(x): \abs{\exists on^* . \set{x}}$: number of blocks above $x$.\footnote{
Action and features names are assigned manually to increase readability, based on the denotations of the involved concepts.
}
\end{itemize}

@@

\begin{itemize}
\item \emph{pick-above-$x$}: \tuple{\neg X, \neg H, n(x) > 0; n(x) \downarrow},
\item \emph{place-aside}: \tuple{\neg X, H; \neg H},
\end{itemize}

@@

\begin{itemize}
\item $X$, $n(x)$, $n(y)$: as in $\Q_{clear}$,
\item $E: handempty$: whether gripper is empty,
\item $XY: [(\exists on^{-1} . \set{x})\sqcap \set{b}]$: whether $x$ is on $y$.
\end{itemize}

\noindent Interestingly, the last feature is exactly a concept-based representation of the actual STRIPS goal $on(x,y)$.
The corresponding action model $A_F$ is:

\begin{itemize}
\item \emph{pick-above-$x$}:  \tuple{\neg X, \neg XY, E, n(x) > 0, n(y)>0; n(x) \downarrow, \neg E  }
\item \emph{pick-above-$y$}:  \tuple{\neg X, \neg XY, E, n(x) = 0, n(y)>0; n(y) \downarrow, \neg E  }
\item \emph{put-$x$-on-$y$}:  \tuple{X, \neg XY, \neg E, n(x)=0, n(y)=0 ; n(y) \uparrow, \neg X, E, XY}
\item \emph{pick-$x$}:  \tuple{;}
\item \emph{put-$x$-on-table}:  \tuple{;}
\item \emph{put-aside-1}:  \tuple{;}
\item \emph{put-aside-2}:  \tuple{;}

\end{itemize}

@@

\alert{ongoing}

This last action will indeed be necessary in case block $x$ is initially being held
but block $y$ is not clear.

@@

\paragraph{Gripper.}
The problem $\Q_{gripper}$ is based on the classical Gripper problem, where a robot with two grippers needs to 
move a number of blocks from one room into another \emph{target} room $x$, which we take to be a parameter of the problem,
each gripper of the robot being able to carry only one ball at a time.
We use the standard STRIPS representation with primitive predicates $at\text{-}robby(l)$, $at\text{-}ball(b, l)$, $free(g)$, $carry(b, g)$ that denote,
respectively, whether the robot (the ball) is at location $l$, whether gripper $g$ is free, and whether gripper $g$ carries ball $b$,
plus unary type predicates $room$, $ball$, and $gripper$.
%
We feed our pipeline with 2 training instances of the problem with only 2 rooms and at most 4 balls.
The resulting abstract action model is based on the following features:

\begin{itemize}
\item $C: \exists carry . \top$ (number of balls carried by the robot), 
\item $X: at\_robby \sqcap \set{x}$ (whether robot in target room),
\item $G: \abs{free}$ (number of empty grippers), and
\item $B: \exists at . \neg \set{x}$ (number of balls not in target room),
\end{itemize}

\noindent and contains the following actions :

\begin{itemize}
\item \emph{drop-ball(s)-at-$x$}: \tuple{C>0, X;  C\downarrow, G \uparrow}
\item \emph{move-to-$x$-half-loaded}: \tuple{\neg X, B=0, C>0, G>0;   X}
\item \emph{move-to-$x$-fully-loaded}: \tuple{\neg X, C>0, G=0;   X}
\item \emph{pick-ball-not-in-$x$}: \tuple{\neg X, B > 0, G > 0;   B \downarrow, G \downarrow, C\uparrow}
\item \emph{drop-$x$}: \tuple{X, C=0, G > 0;   \neg X}
\end{itemize}
\noindent \alert{Action model to be checked}

The model is not complete, as it e.g. features no action to pick a ball from room $x$,
but is goal-oriented as a result of using the SAT theory $T_G$, in that it features all of the actions
necessary to achieve the goal.
This can be seen e.g. in the fact that the action to move \emph{move-to-$x$-half-loaded}
can only be performed when no more balls remain in the other room.
Additionally, the model is \emph{provably correct for any gripper instance with an arbitrary number of balls, 
rooms and grippers}. CHECK !!

@@

% Action 1:
%         PRE: ncarried > 0, robot-at-B
%         EFFS: DEC ncarried, INC nfree-grippers
% 
% Action 2:
%         PRE: NOT robot-at-B, nballs-A = 0, ncarried > 0, nfree-grippers > 0
%         EFFS: robot-at-B
% 
% Action 3:
%         PRE: NOT robot-at-B, ncarried > 0, nfree-grippers = 0
%         EFFS: robot-at-B
% 
% Action 4:
%         PRE: NOT robot-at-B, nballs-A > 0, nfree-grippers > 0
%         EFFS: DEC nballs-A, DEC nfree-grippers, INC ncarried
% 
% Action 5:
%         PRE: ncarried = 0, nfree-grippers > 0, robot-at-B
%         EFFS: NOT robot-at-B

\paragraph{Collecting rewards in a maze.}
The objective in the problem $\Q_{reward}$ is to move in an $n \times m$ grid to pick up certain
rewards that are spread out in certain grid cells, while avoiding other cells which are considered as \emph{blocked}.
This example is from recent work in symbolic deep
reinforcement learning, where there is a penalty associated for
stepping in blocked cells instead \cite{garnelo2016towards}.
%
We represent the instances $P$ with primitive predicates $reward(c)$, $at(c)$, $blocked(c)$ and $adjacent(c_1, c_2)$ that denote
respectively the position of the rewards, of the agent, of the blocked cells, and the grid topology.
We use two single instances of the problem with size $5 \times 5$ and $4 \times 4$ and different distributions of blocks and rewards.
We learn a set with 2 features:

- $R: \abs{reward}$ (number of remaining rewards), and

- $D: dist(at, adjacent:(\neg blocked), reward)$ (distance from current cell to closest cell with reward, traversing adjacent, unblocked cells only).

\noindent The action model has also 2 actions:

- \emph{pick-reward}: \tuple{D=0, R > 0;  R \downarrow, D \uparrow}

- \emph{move-towards-closest-reward}: \tuple{R>0, D>0; D \downarrow}

The model is not complete ... 
...

The policy ... 
... EXPAND A BIT - TOO TIRED NOW.
Discuss suboptimality of the policy.
Perhaps discuss D=$\infty$ when no reward?

>>>>

\begin{itemize}
\item $H: \abs{holding}$, whether any block is being held,
\item $X: \abs{holding \sqcap \set{x}}$, whether block $x$ is being held,
\item $n(x): \abs{\exists on^* . \set{x}}$, number of blocks above $x$.\footnote{
Action and features names are assigned manually to increase readability, based on the denotations of the involved concepts.
}
\end{itemize}

@@

\begin{itemize}
\item \emph{pick-above-$x$}: \hspace*{\fill} \tuple{\neg X, \neg H, n(x) > 0; X, n(x) \downarrow},
\item \emph{place-aside}: \hspace*{\fill} \tuple{\neg X, H; \neg H},
\end{itemize}

@@

\begin{itemize}
\item $X$, $n(x)$, $n(y)$, as in $\Q_{clear}$,
\item $E: handempty$, whether gripper is empty,
\item $XY: [(\exists on^{-1} . \set{x})\sqcap \set{b}]$, whether $x$ is on $y$.
\end{itemize}

\noindent Interestingly, the last feature is a concept-based representation of the actual STRIPS goal $on(x,y)$.
The corresponding action model $A_F$ is:

\begin{itemize}
% ATM I use two lines, as doesn't fit in one :-(
\item \emph{pick-above-$x$}: \\\hspace*{\fill}  \tuple{\neg X, \neg XY, E, n(x) > 0, n(y)>0; n(x) \downarrow, \neg E  }

\item \emph{pick-above-$y$}: \\\hspace*{\fill}  \tuple{\neg X, \neg XY, E, n(x) = 0, n(y)>0; n(y) \downarrow, \neg E  }

\item \emph{put-$x$-on-$y$}: \\\hspace*{\fill} \tuple{X, \neg XY, \neg E, n(x)=n(y)=0 ; n(y) \uparrow, \neg X, E, XY}

\item \emph{pick-$x$}:         \hspace*{\fill}  \tuple{\neg X, \neg XY, E, n(x)=0, n(y)=0; E, X}

\item \emph{put-$x$-table}:                     \tuple{\neg E, \neg XY, X, n(x)=0, n(y) > 0; \neg X, E}

\item \emph{put-aside-1}:      \hspace*{\fill} \tuple{\neg E, \neg X, \neg XY, n(x) = 0; E}

\item \emph{put-aside-2}:      \hspace*{\fill} \tuple{\neg E, \neg X, \neg XY, n(x) > 0, n(y) > 0; E}
\end{itemize}

@@

Action \emph{put-$x$-table} will indeed be necessary only in case block $x$ starts being held
but block $y$ is not clear.
%
The two last \emph{put-aside} actions show that when using the goal-oriented theory $T_G(\S,\F)$
it is not always possible to completely merge similar actions.
This is because some states (in this case, any state with $n(x) > 0 \land n(y) = 0$ and some block $b \not\in \set{x,y}$ being held)
by chance did not appear on the optimal plans computed, which makes the resulting $A_F$ not complete in general.
This limitation could be circumvented either by manually providing an additional 
instance fulfilling that condition, or, more robustly, by using a larger set of optimal plans instead of just one.

Interestingly, action \emph{pick-$x$} illustrates a different type of incompleteness, as it will
only pick block $x$ whenever block $y$ is clear. No alternative optimal plan will possibly fix this,
since (provided $x$ and $y$ are initially not in the same tower) no optimal plan for $on(x,y)$ will pick block $x$ before $y$ is clear.
This incompleteness however will make the computation of the policies easier and the resulting policies likely more effective.

The policy ...
\alert{Display policy here}

\paragraph{Gripper.}
The problem $\Q_{gripper}$ is based on the classical Gripper problem, where a robot with two grippers needs to 
move a number of blocks from one room into another \emph{target} room $x$, which we take to be a parameter of the problem,
each gripper of the robot being able to carry only one ball at a time.
We use the standard STRIPS representation with primitive predicates $at\text{-}robby(l)$, $at\text{-}ball(b, l)$, $free(g)$, $carry(b, g)$ that denote,
respectively, whether the robot (the ball) is at location $l$, whether gripper $g$ is free, and whether gripper $g$ carries ball $b$,
plus unary type predicates $room$, $ball$, and $gripper$.
%
Using 2 training instances $P$ with only 2 rooms and at most 4 balls, we learn an abstract state space
based on the following features:

\begin{itemize}
\item $C: \abs{\exists carry . \top}$, number of balls carried by the robot, 
\item $X: at\_robby \sqcap \set{x}$, whether robot in target room,
\item $G: \abs{free}$, number of empty grippers, and
\item $B: \abs{\exists at . \neg \set{x}}$, number of balls not in target room,
\end{itemize}

\noindent and an abstract action model $A_F$:

\begin{itemize}
\item \emph{drop-ball-at-$x$}: \hspace*{\fill} \tuple{C>0, X;  C\downarrow, G \uparrow}
\item \emph{move-to-$x$-fully-loaded}: \hspace*{\fill} \tuple{\neg X, C>0, G=0;   X}
\item \emph{move-to-$x$-half-loaded}: \hspace*{\fill} \tuple{\neg X, B=0, C>0, G>0;   X}
\item \emph{pick-ball-not-in-$x$}: \hspace*{\fill} \tuple{\neg X, B > 0, G > 0;   B \downarrow, G \downarrow, C\uparrow}
\item \emph{leave-$x$}: \hspace*{\fill} \tuple{X, C=0, G > 0;   \neg X}
\end{itemize}

The marking of only states in optimal plans as \emph{goal-relevant} introduces again a bias
that can be seen e.g. in the fact that the robot is not allowed to leave room $x$ if carrying any ball,
or that \emph{move-to-$x$-half-loaded} is only allowed in the last move, when no more balls remain in the other room.
Although allowing moves without full load before that would help minimizing the action model and achieving completeness, it would certainly \emph{not}
help obtaining better policies.
%
Despite this incompleteness, the action model is \emph{provably correct for any $\Q_{gripper}$ instance with an arbitrary number of balls, 
rooms and grippers}.

@@

% Action 1:
%         PRE: ncarried > 0, robot-at-B
%         EFFS: DEC ncarried, INC nfree-grippers
% 
% Action 2: [move-to-x-half-loaded]
%         PRE: NOT robot-at-B, nballs-A = 0, ncarried > 0, nfree-grippers > 0
%         EFFS: robot-at-B
% 
% Action 3: [move-to-x-fully-loaded]
%         PRE: NOT robot-at-B, ncarried > 0, nfree-grippers = 0
%         EFFS: robot-at-B
% 
% Action 4:
%         PRE: NOT robot-at-B, nballs-A > 0, nfree-grippers > 0
%         EFFS: DEC nballs-A, DEC nfree-grippers, INC ncarried
% 
% Action 5:
%         PRE: ncarried = 0, nfree-grippers > 0, robot-at-B
%         EFFS: NOT robot-at-B

\paragraph{Collecting rewards in a maze.}
The objective in the problem $\Q_{reward}$ is to move in an $n \times m$ grid to pick up certain
rewards that are spread out in certain grid cells, while avoiding other cells which are considered as \emph{blocked}.
This example is from recent work in symbolic deep
reinforcement learning, where there is a penalty associated for
stepping in blocked cells instead \cite{garnelo2016towards}.
%
We represent the instances $P$ with primitive predicates $reward(c)$, $at(c)$, $blocked(c)$ and $adjacent(c_1, c_2)$ that denote
respectively the position of the rewards, of the agent, of the blocked cells, and the grid topology.
This is the only problem where we test the $\textit{dist}(C_1,R:C,C_2)$ min-distance features introduced before.
We use two single training instances with grid sizes $5 \times 5$ and $4 \times 4$ and different distributions of blocks and rewards,
and learn an abstraction with 2 features:

\begin{itemize}
\item $R: \abs{reward}$, number of remaining rewards, and
\item $D: dist(at, adjacent:(\neg blocked), reward)$, distance from current cell to closest cell with reward, traversing adjacent, unblocked cells only.
\end{itemize}

\noindent The action model has the following 2 actions:

\begin{itemize}
\item \emph{pick-reward}: \hspace*{\fill} \tuple{D=0, R > 0;  R \downarrow, D \uparrow}
\item \emph{move-towards-closest-reward}: \hspace*{\fill} \tuple{R>0, D>0; D \downarrow}
\end{itemize}

The model is very simple, and can be shown to solve any $\Q_{reward}$ instance irrespective of 
grid size and number and location of rewards and cell blocks, \emph{as long as the instance is solvable},
i.e. all rewards are reachable.
\alert{Interesting point here: how does our approach fare with unsolvability? And with problems with dead-end?}

@@

Discuss suboptimality of the policy?
